---
title: "Risk Managemen"
authot: "Nikita Tatarinov"
output: html_document
date: "2023-10-26"
---

## 6.0. Data

GT ID (seed): 903914813

### 6.0.1. DSF
```{r}
library(data.table)
dsf <- fread("./data/dsf.csv", select = c("PERMNO", "date", "RET", "vwretd"))
dsf$RET[dsf$RET == "B" | dsf$RET == "C"] <- NA
dsf$RET <- as.numeric(dsf$RET)
dsf <- na.omit(dsf)
summary(dsf)
```

### 6.0.2. Creating individual porfolio and date periods
```{r}
set.seed(903914813)
first_year <- sample(1981:2000, 1, replace = FALSE) # boundaries are included
first_date <- as.Date(paste0(first_year, "-01-01"), format = "%Y-%m-%d")
first_year_end_date <- as.Date(paste0(first_year, "-12-31"), format = "%Y-%m-%d")
last_date <- as.Date(paste0((first_year+9), "-12-31"), format = "%Y-%m-%d")

first_year_dsf <- dsf[(dsf$date >= first_date) & (dsf$date <= first_year_end_date)]
unique_stocks <- unique(first_year_dsf$PERMNO)
num_unique_stocks <- length(unique_stocks)
stock_indices <- sample(1:num_unique_stocks, 250, replace = FALSE)
stocks <- unique_stocks[stock_indices]
dsf <- dsf[dsf$PERMNO %in% stocks]

dsf_2001_2011 <- dsf[(dsf$date >= as.Date("2001-01-01")) & (dsf$date <= as.Date("2011-12-31"))]
dsf_2012_2022 <- dsf[(dsf$date >= as.Date("2012-01-01")) & (dsf$date <= as.Date("2022-12-31"))]
dsf <- dsf[(dsf$date >= first_date) & (dsf$date <= last_date)]

summary(dsf)
summary(dsf_2001_2011)
summary(dsf_2012_2022)
```

### 6.0.3. Risk-free rate
```{r}
library(data.table)
rf <- fread("./data/F-F_Research_Data_Factors_daily.CSV", skip=4, select = c(1, 5))
colnames(rf)[which(colnames(rf) == "V1")] <- "date"
rf$date <- as.Date(as.character(rf$date), format = "%Y%m%d")
summary(rf)
```

### 6.0.4. Merging and getting excess returns
```{r}
convert_returns_to_excess <- function(df) {
  df <- merge(df, rf, by = "date")
  df$RET <- df$RET - df$RF
  df$vwretd <- df$vwretd - df$RF
  colnames(df)[which(colnames(df) == "RET")] <- "EXCRET"
  colnames(df)[which(colnames(df) == "vwretd")] <- "EXCMKT"
  df$RF <- NULL
  return(df)
}

dsf <- convert_returns_to_excess(dsf)
dsf_2001_2011 <- convert_returns_to_excess(dsf_2001_2011)
dsf_2012_2022 <- convert_returns_to_excess(dsf_2012_2022)

summary(dsf)
summary(dsf_2001_2011)
summary(dsf_2012_2022)
```

## 6.1. Semi Beta, Downside Beta, Co-skewness and Tail Risk

```{r}
library(dplyr)
dsf <- dsf %>% mutate(year_month = format(date, "%Y-%m"))
dsf$year_month <- as.Date(paste0(dsf$year_month, "-01"))

calculate_f <- function(df) {
  mean_mkt = mean(dsf$EXCMKT)
  df <- df %>% mutate(f_tk = EXCMKT - mean_mkt)
  df <- df %>% mutate(f_tk_plus = ifelse(f_tk > 0, f_tk, 0))
  df <- df %>% mutate(f_tk_minus = ifelse(f_tk < 0, f_tk, 0))
  
  num_cols <- length(unique(df$PERMNO))
  df <- df %>% mutate(beta_denominator = f_tk * f_tk)
  df$beta_denominator <- sum(df$beta_denominator) / num_cols
  df <- df %>% mutate(beta_plus_denominator = f_tk_plus * f_tk_plus)
  df$beta_plus_denominator <- sum(df$beta_plus_denominator) / num_cols
  df <- df %>% mutate(beta_minus_denominator = f_tk_minus * f_tk_minus)
  df$beta_minus_denominator <- sum(df$beta_minus_denominator) / num_cols
  return(df)
}
dsf <- dsf %>% group_by(year_month) %>% do(calculate_f(.)) %>% ungroup()

calculate_r_betas_coskew_cokurt <- function(df) {
  mean_ret <- mean(dsf$EXCRET)
  df <- df %>% mutate(r_tki = EXCRET - mean_ret)
  df <- df %>% mutate(r_tki_plus = ifelse(r_tki > 0, r_tki, 0))
  df <- df %>% mutate(r_tki_minus = ifelse(r_tki < 0, r_tki, 0))
  
  df <- df %>% mutate(beta_N_numenator = r_tki_minus * f_tk_minus)
  df$beta_N_numenator <- sum(df$beta_N_numenator)
  df <- df %>% mutate(beta_N = beta_N_numenator / beta_denominator)
  
  df <- df %>% mutate(beta_P_numenator = r_tki_plus * f_tk_plus)
  df$beta_P_numenator <- sum(df$beta_P_numenator)
  df <- df %>% mutate(beta_P = beta_P_numenator / beta_denominator)
  
  df <- df %>% mutate(beta_M_minus_numenator = r_tki_plus * f_tk_minus)
  df$beta_M_minus_numenator <- sum(df$beta_M_minus_numenator)
  df <- df %>% mutate(beta_M_minus = beta_M_minus_numenator / beta_denominator)
  
  df <- df %>% mutate(beta_M_plus_numenator = r_tki_minus * f_tk_plus)
  df$beta_M_plus_numenator <- sum(df$beta_M_plus_numenator)
  df <- df %>% mutate(beta_M_plus = beta_M_plus_numenator / beta_denominator)
  
  df <- df %>% mutate(beta = beta_N + beta_P - beta_M_minus - beta_M_plus)
  
  df <- df %>% mutate(beta_minus_numenator = r_tki * f_tk_minus)
  df$beta_minus_numenator <- sum(df$beta_minus_numenator)
  df <- df %>% mutate(beta_minus = beta_minus_numenator / beta_minus_denominator)
  
  df <- df %>% mutate(beta_plus_numenator = r_tki * f_tk_plus)
  df$beta_plus_numenator <- sum(df$beta_plus_numenator)
  df <- df %>% mutate(beta_plus = beta_plus_numenator / beta_plus_denominator)
  
  m <- length(df$EXCRET)
  
  df <- df %>% mutate(coskew_numenator = r_tki * f_tk * f_tk)
  df$coskew_numenator <- sum(df$coskew_numenator) / m
  df <- df %>% mutate(coskew_denominator_1 = r_tki * r_tki)
  df$coskew_denominator_1 <- sqrt(sum(df$coskew_denominator_1) / m)
  df <- df %>% mutate(coskew_denominator_2 = f_tk * f_tk)
  df$coskew_denominator_2 <- sum(df$coskew_denominator_2) / m
  df <- df %>% mutate(coskew = coskew_numenator / (coskew_denominator_1 * coskew_denominator_2))
  
  df <- df %>% mutate(cokurt_numenator = r_tki * f_tk * f_tk * f_tk)
  df$cokurt_numenator <- sum(df$cokurt_numenator) / m
  df <- df %>% mutate(cokurt_denominator_1 = r_tki * r_tki)
  df$cokurt_denominator_1 <- sqrt(sum(df$cokurt_denominator_1) / m)
  df <- df %>% mutate(cokurt_denominator_2 = f_tk * f_tk)
  df$cokurt_denominator_2 <- (sum(df$cokurt_denominator_2) / m)^(3/2)
  df <- df %>% mutate(cokurt = cokurt_numenator / (cokurt_denominator_1 * cokurt_denominator_2))
  
  return(df)
}
dsf <- dsf %>% group_by(PERMNO, year_month) %>% do(calculate_r_betas_coskew_cokurt(.)) %>% ungroup()

dsf$f_tk <- NULL
dsf$f_tk_plus <- NULL
dsf$f_tk_minus <- NULL
dsf$beta_denominator <- NULL
dsf$beta_plus_denominator <- NULL
dsf$beta_minus_denominator <- NULL
dsf$r_tki <- NULL
dsf$r_tki_plus <- NULL
dsf$r_tki_minus <- NULL
dsf$beta_N_numenator <- NULL
dsf$beta_P_numenator <- NULL
dsf$beta_M_minus_numenator <- NULL
dsf$beta_M_plus_numenator <- NULL
dsf$beta_minus_numenator <- NULL
dsf$beta_plus_numenator <- NULL
dsf$coskew_numenator <- NULL
dsf$coskew_denominator_1 <- NULL
dsf$coskew_denominator_2 <- NULL
dsf$cokurt_numenator <- NULL
dsf$cokurt_denominator_1 <- NULL
dsf$cokurt_denominator_2 <- NULL

summary(dsf)
```

As we can see, more than overall values seem to be adequate, but there are outliers which make betas more than 10. They were observed in previous assignments too and winsorized. According to coskew, most of companies' returns seem to be skewed similarly to the market. But as for cokurt, tails of companies' return distribution seems to be more fat-tailed, which illustrates that companies are generally more risky.

## 6.2. Value-at-Risk (VaR) and Expected Shortfall (ES)

```{r}
library(dplyr)

calculate_es <- function(EXCRET) {
  value_at_risk = quantile(EXCRET, probs=0.05)
  return(mean(EXCRET[EXCRET <= value_at_risk]))
}

calculate_var_df <- function(df) {
  VAR_df <- df %>% group_by(PERMNO) %>% summarize(value_at_risk = quantile(EXCRET, probs=0.05)) %>% ungroup()
  VAR_df <- VAR_df %>% mutate(value_at_risk_dollar = 1000000 * (1 - value_at_risk))
  VAR_df$ES <- (df %>% group_by(PERMNO) %>% summarize(ES = calculate_es(EXCRET)) %>% ungroup())$ES
  return(VAR_df)
}

print("For individual portfolio")
VAR_dsf = calculate_var_df(dsf)
summary(VAR_dsf)
head(VAR_dsf, 20)

print("For 2001-2011")
VAR_dsf_2001_2011 = calculate_var_df(dsf_2001_2011)
summary(VAR_dsf_2001_2011)
head(VAR_dsf_2001_2011, 10)
```

As we can see, there are now marked alterations in terms of VaR and related characteristics. We can observe narrowing of the characteristics in terms of minimum and maximum values, but overall the distribtions seem to be quite similar.

## 6.3. Volatility modeling

```{r}
unique_stocks_2001 <- unique(dsf_2001_2011$PERMNO)
unique_stocks_2001 <- unique_stocks_2001[unique_stocks_2001 %in% stocks]
modeling_stock_indices <- sample(1:length(unique_stocks_2001), 5, replace = FALSE)
modeling_stocks <- unique_stocks_2001[modeling_stock_indices]

sorted_dsf <- dsf[dsf$PERMNO %in% modeling_stocks,]
sorted_dsf <- sorted_dsf[order(sorted_dsf$date), ]
sorted_dsf <- sorted_dsf %>% mutate(EXCRET_sqr = EXCRET * EXCRET)
variance_df <- sorted_dsf %>% group_by(PERMNO) %>% summarize(variance = var(EXCRET), ret_sqr = mean(EXCRET_sqr)) %>% ungroup()
variance_df <- variance_df %>% mutate(t = 0)

sorted_dsf <- dsf_2001_2011[dsf_2001_2011$PERMNO %in% modeling_stocks,]
sorted_dsf <- sorted_dsf[order(sorted_dsf$date), ]
sorted_dsf <- sorted_dsf %>% mutate(EXCRET_sqr = EXCRET * EXCRET)
variance_df <- variance_df %>% mutate(date = sorted_dsf$date[1])
summary(variance_df)
```

```{r}
lambda = 0.94
dsf_by_date <- split(sorted_dsf, sorted_dsf$date)
for (i in 1:length(dsf_by_date)) {
  for (j in 1:length(dsf_by_date[[i]]$PERMNO)) {
    if (dsf_by_date[[i]]$PERMNO[j] %in% stocks) {
      prev_row <- variance_df[(variance_df$PERMNO == as.numeric(dsf_by_date[[i]]$PERMNO[j])) & (variance_df$t == i - 1), ]
      k <- i - 2
      while (nrow(prev_row) != 1) {
        prev_row <- variance_df[(variance_df$PERMNO == as.numeric(dsf_by_date[[i]]$PERMNO[j])) & (variance_df$t == k), ]
        k <- k - 1
      }
      prev_variance <- prev_row$variance
      prev_ret_sqr <- prev_row$ret_sqr
      new_variance <- lambda * prev_variance + (1 - lambda) * prev_ret_sqr
      new_ret_sqr <- as.numeric(dsf_by_date[[i]]$EXCRET_sqr[j])
      new_row <- data.frame(PERMNO = as.numeric(dsf_by_date[[i]]$PERMNO[j]), 
                            variance = new_variance, 
                            ret_sqr = new_ret_sqr, 
                            t = i,
                            date = dsf_by_date[[i]]$date[j])
      variance_df <- rbind(variance_df, new_row)
    }
  }
}

variance_df <- variance_df[order(variance_df$t, variance_df$PERMNO), ]
summary(variance_df)
```


```{r}
library(ggplot2)
to_plot <- variance_df[variance_df$PERMNO == modeling_stocks[[1]], ]
ggplot(to_plot, aes(x = as.Date(date), y = variance)) +
  geom_line() +
  labs(x = "Date", y = "Variance") +
  ggtitle(modeling_stocks[[1]])
```

```{r}
library(ggplot2)
to_plot <- variance_df[variance_df$PERMNO == modeling_stocks[[2]], ]
ggplot(to_plot, aes(x = as.Date(date), y = variance)) +
  geom_line() +
  labs(x = "Date", y = "Variance") +
  ggtitle(modeling_stocks[[2]])
```

```{r}
library(ggplot2)
to_plot <- variance_df[variance_df$PERMNO == modeling_stocks[[3]], ]
ggplot(to_plot, aes(x = as.Date(date), y = variance)) +
  geom_line() +
  labs(x = "Date", y = "Variance") +
  ggtitle(modeling_stocks[[3]])
```

```{r}
library(ggplot2)
to_plot <- variance_df[variance_df$PERMNO == modeling_stocks[[4]], ]
ggplot(to_plot, aes(x = as.Date(date), y = variance)) +
  geom_line() +
  labs(x = "Date", y = "Variance") +
  ggtitle(modeling_stocks[[4]])
```

```{r}
library(ggplot2)
to_plot <- variance_df[variance_df$PERMNO == modeling_stocks[[5]], ]
ggplot(to_plot, aes(x = as.Date(date), y = variance)) +
  geom_line() +
  labs(x = "Date", y = "Variance") +
  ggtitle(modeling_stocks[[5]])
```

Generally, variance seems to be quite volatile. However, clear peaks are observed for all of the companies in 2001 and 2009, which may represent crisis periods. Moreover, the last company just stopped existing in 2001.

## GARCH

```{r}
sorted_dsf <- dsf[dsf$PERMNO %in% modeling_stocks, c("PERMNO", "date", "EXCRET")]
colnames(sorted_dsf)[which(colnames(sorted_dsf) == "EXCRET")] <- "returns"
sorted_dsf <- sorted_dsf[order(sorted_dsf$date), ]
summary(sorted_dsf)

garch_df <- data.frame(PERMNO = modeling_stocks, omega = 0, beta = 0, alpha = 0)
head(garch_df)
```

```{r}
library("rugarch")
library("lubridate")

sorted_dsf$date <- as.POSIXct(sorted_dsf$date)
spec <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(1, 1)), distribution.model = "sstd")
for (stock in modeling_stocks) {
  model <- ugarchfit(spec, data = sorted_dsf[sorted_dsf$PERMNO == stock, c("date", "returns")])
  summary(model)
  coefficients <- model@fit$coef
  garch_df[garch_df$PERMNO == as.numeric(stock), ]$omega = model@fit$coef[["omega"]]
  garch_df[garch_df$PERMNO == as.numeric(stock), ]$beta = model@fit$coef[["beta1"]]
  garch_df[garch_df$PERMNO == as.numeric(stock), ]$alpha = model@fit$coef[["alpha1"]]
}

head(garch_df)
```

```{r}
sorted_dsf <- sorted_dsf %>% mutate(EXCRET_sqr = returns * returns)
garch_variance_df <- sorted_dsf %>% group_by(PERMNO) %>% summarize(variance = var(returns), ret_sqr = mean(EXCRET_sqr)) %>% ungroup()
garch_variance_df <- garch_variance_df %>% mutate(t = 0)

sorted_dsf <- dsf_2012_2022[dsf_2012_2022$PERMNO %in% modeling_stocks, c("PERMNO", "date", "EXCRET")]
sorted_dsf <- sorted_dsf[order(sorted_dsf$date), ]
sorted_dsf <- sorted_dsf %>% mutate(EXCRET_sqr = EXCRET * EXCRET)
garch_variance_df <- garch_variance_df %>% mutate(date = sorted_dsf$date[1])
summary(garch_variance_df)
```

```{r}
dsf_by_date <- split(sorted_dsf, sorted_dsf$date)
for (i in 1:length(dsf_by_date)) {
  for (j in 1:length(dsf_by_date[[i]]$PERMNO)) {
    if (dsf_by_date[[i]]$PERMNO[j] %in% stocks) {
      prev_row <- garch_variance_df[(garch_variance_df$PERMNO == as.numeric(dsf_by_date[[i]]$PERMNO[j])) & 
                                      (garch_variance_df$t == i - 1), ]
      k <- i - 2
      while (nrow(prev_row) != 1) {
        prev_row <- variance_df[(garch_variance_df$PERMNO == as.numeric(dsf_by_date[[i]]$PERMNO[j])) & 
                                  (garch_variance_df$t == k), ]
        k <- k - 1
      }
      prev_variance <- prev_row$variance
      prev_ret_sqr <- prev_row$ret_sqr
      garch_coeffs <- garch_df[garch_df$PERMNO == dsf_by_date[[i]]$PERMNO[j], ]
      new_variance <- garch_coeffs$omega + garch_coeffs$beta * prev_variance + garch_coeffs$alpha * prev_ret_sqr
      new_ret_sqr <- as.numeric(dsf_by_date[[i]]$EXCRET_sqr[j])
      new_row <- data.frame(PERMNO = as.numeric(dsf_by_date[[i]]$PERMNO[j]), 
                            variance = new_variance, 
                            ret_sqr = new_ret_sqr, 
                            t = i,
                            date = dsf_by_date[[i]]$date[j])
      garch_variance_df <- rbind(garch_variance_df, new_row)
    }
  }
}

garch_variance_df <- garch_variance_df[order(garch_variance_df$t, garch_variance_df$PERMNO), ]
summary(garch_variance_df)
```

```{r}
library(ggplot2)
to_plot <- garch_variance_df[garch_variance_df$PERMNO == modeling_stocks[[1]], ]
ggplot(to_plot, aes(x = as.Date(date), y = variance)) +
  geom_line() +
  labs(x = "Date", y = "Variance") +
  ggtitle(modeling_stocks[[1]])
```

```{r}
library(ggplot2)
to_plot <- garch_variance_df[garch_variance_df$PERMNO == modeling_stocks[[2]], ]
ggplot(to_plot, aes(x = as.Date(date), y = variance)) +
  geom_line() +
  labs(x = "Date", y = "Variance") +
  ggtitle(modeling_stocks[[2]])
```

```{r}
library(ggplot2)
to_plot <- garch_variance_df[garch_variance_df$PERMNO == modeling_stocks[[3]], ]
ggplot(to_plot, aes(x = as.Date(date), y = variance)) +
  geom_line() +
  labs(x = "Date", y = "Variance") +
  ggtitle(modeling_stocks[[3]])
```

```{r}
library(ggplot2)
to_plot <- garch_variance_df[garch_variance_df$PERMNO == modeling_stocks[[4]], ]
ggplot(to_plot, aes(x = as.Date(date), y = variance)) +
  geom_line() +
  labs(x = "Date", y = "Variance") +
  ggtitle(modeling_stocks[[4]])
```

```{r}
library(ggplot2)
to_plot <- garch_variance_df[garch_variance_df$PERMNO == modeling_stocks[[5]], ]
ggplot(to_plot, aes(x = as.Date(date), y = variance)) +
  geom_line() +
  labs(x = "Date", y = "Variance") +
  ggtitle(modeling_stocks[[5]])
```

Again, a distinct peak is seen in 2020, during COVID, which should prove the influence of financial crisis on volatility. What is particularly interesting, only 3 out of 5 companies ceased to exist in 2010s.